---
title: "Getting Started with SymbolizeR"
output: 
  rmarkdown::html_vignette:
    css: vignette-theme.css
    includes:
      after_body: vignette-extras.html
vignette: >
  %\VignetteIndexEntry{Getting Started with SymbolizeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
```


SymbolizeR is a lightweight **symbolic probability engine** for R. It derives expectations, variances, and covariances analytically—returning formulas, not numbers.

```{r setup}
library(SymbolizeR)
```

---

## The Capital Letter Convention

SymbolizeR infers variable types from naming conventions:

| First Character | Type | Examples |
|-----------------|------|----------|
| **Uppercase** (A-Z) | Random Variable | `X`, `Y`, `Normal` |
| **Lowercase** (a-z) | Constant | `a`, `mu`, `sigma` |
| **Numeric** | Number | `1`, `2.5`, `42` |

No declarations needed—just use intuitive names:

```{r convention}
E(X)      # X is a random variable
E(mu)     # mu is a constant  
E(5)      # 5 is a number
```

---

## Symbolic Expectation: `E()`

The `E()` function applies **linearity of expectation** symbolically.

### Core Properties

$$E[aX + b] = aE[X] + b \qquad E[X + Y] = E[X] + E[Y]$$

```{r expectation}
# Constants factor out
E(a * X)

# Distributes over sums
E(X + Y)

# Linear combinations
E(a * X + b)

# Numeric coefficients
E(2 * X + 3)
```

### Products of Random Variables

Without independence assumptions, products cannot be simplified:

```{r product}
E(X * Y)
```

**Declare independence** to factor products:

```{r independence}
assume.independent(X, Y)
E(X * Y)  # Now factors to E(X) * E(Y)
clear.independence()  # Reset for subsequent examples
```

---

## Symbolic Variance: `Var()`

Computes variance using: $\text{Var}(X) = E[X^2] - E[X]^2$

```{r variance}
Var(X)

Var(a * X)
```

---

## Symbolic Covariance: `Cov()`

Computes covariance using: $\text{Cov}(X, Y) = E[XY] - E[X]E[Y]$

```{r covariance}
Cov(X, Y)
```

---

## Limitations

SymbolizeR performs **symbolic manipulation only**:

- ✗ No numerical evaluation
- ✗ Complex expressions returned as-is

```{r limitations}
E(1/X)    # Cannot simplify

E(X^Y)    # Non-linear in RV
```

---

## Next Steps

| Topic | Learn About |
|-------|-------------|
| [**Distributions**](distributions.html) | Automatic moment substitution with `define()` |
| [**Derivations**](derivations.html) | Step-by-step work with `derive.E()`, `derive.Var()` |
| [**Calculus**](calculus.html) | Symbolic differentiation and integration |
| [**Independence & LaTeX**](independence-latex.html) | Declare independence, export to LaTeX |
| [**Package Internals**](internals.html) | How SymbolizeR works under the hood |

<div class="vignette-nav">
  <div></div>
  <a href="distributions.html" class="nav-item next">
    <span class="nav-label">Next</span>
    <span class="nav-title">2. Distributions</span>
  </a>
</div>

