---
title: "Step-by-Step Derivations"
output: 
  html_document:
    template: template.html
    self_contained: false
    toc: false
    theme: null
    highlight: pygments
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Step-by-Step Derivations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
```


```{r setup}
library(SymbolizeR)
```

SymbolizeR's derivation functions show **every step** of the calculation—invaluable for learning, teaching, and verification.

---

## `derive.E()` — Expectation Derivation

```{r derive-e}
derive.E(a * X + b)
```

### With Defined Distributions

```{r derive-e-defined}
clear.definitions()
define(X ~ Normal(mu, sigma))

derive.E(a * X + b)

clear.definitions()
```

### Linear Combinations

```{r derive-e-linear}
derive.E(X + Y + Z)
```

---

## `derive.Var()` — Variance Derivation

Shows complete variance calculation using $\text{Var}(X) = E[X^2] - E[X]^2$:

```{r derive-var}
derive.Var(X)
```

### Linear Transformations

See polynomial expansion for $\text{Var}(aX + b)$:

```{r derive-var-linear}
derive.Var(a * X + b)
```

### With Defined Distribution

```{r derive-var-defined}
clear.definitions()
define(X ~ Normal(mu, sigma))

derive.Var(X)

clear.definitions()
```

---

## `derive.Cov()` — Covariance Derivation

Shows covariance using $\text{Cov}(X, Y) = E[XY] - E[X]E[Y]$:

```{r derive-cov}
derive.Cov(X, Y)
```

---

## Understanding the Output

Each derivation returns a `derivation` object:

| Component | Description |
derive.Cov(X, X + Y)
```

It correctly identifies that $\text{Cov}(X, X)$ becomes $\text{Var}(X)$.

---

## Why Use This?

1.  **Debugging Math**: If your hand-derived result doesn't match your simulation, use this to find the line where you made a mistake.
2.  **Learning**: It helps build intuition for *why* certain terms cancel out.
3.  **Documentation**: You can copy-paste these derivations directly into your homework or reports.

<div class="page-nav">
  <a href="distributions.html" class="page-nav-item prev">
    <span class="page-nav-label">Previous</span>
    <span class="page-nav-title">← Distributions</span>
  </a>
  <a href="moments.html" class="page-nav-item next">
    <span class="page-nav-label">Next</span>
    <span class="page-nav-title">Moments & Statistics →</span>
  </a>
</div>

