---
title: "Getting Started with SymbolizeR"
output: 
  html_document:
    template: template.html
    self_contained: false
    toc: false
    theme: null
    highlight: pygments
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Getting Started with SymbolizeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
```


Most R packages love numbers. Give them data, they give you separate statistics. **SymbolizeR** is different—it cares about the *formulas* themselves. It lets you write statistical expressions in plain R code and see how they simplify, without needing to plug in a single data point first.

It's essentially a symbolic math engine, but one that speaks the language of probability theory natively.

```{r setup}
library(SymbolizeR)
```

---

## The "Textbook" Convention

When we built SymbolizeR, we didn't want to force you to manually declare every single variable before you used it. That gets tedious fast. 

Instead, we borrowed a convention you've probably seen in every probability textbook ever written: **Capital letters are Random Variables, and lowercase letters are constants.**

It's a simple heuristic, but it makes the code feel almost like handwriting:

| Variable | Type | Why? |
|----------|------|------|
| `X`, `Y`, `Theta` | **Random Variable** | Starts with uppercase (A-Z) |
| `mu`, `sigma`, `a` | **Constant** | Starts with lowercase (a-z) |
| `42`, `1.5` | **Number** | It's just a number |

You don't need to register these ahead of time. Just start typing:

```{r convention}
E(X)      # R sees a capital 'X', so it treats it as random
E(mu)     # R sees a lowercase 'mu', so it knows it's constant
E(5)      # Numbers are just numbers
```

---

## Linearity is Built In

The core of the package is the `E()` function. It's not just a wrapper; it's an engine that understands the **Linearity of Expectation**. It knows that constants can be pulled out and sums can be split apart.

This allows you to simplify expressions symbolically:

```{r expectation}
# Expanding a linear transformation
E(a * X + b)

# Mixing constants and variables
E(2 * X + 3 * Y)
```

Notice what happened there? SymbolizeR automatically applied the rules $E[aX + b] = aE[X] + b$ and $E[X + Y] = E[X] + E[Y]$. It does the algebra for you.

### What about multiplication?

Multiplication is trickier. If you have $E[XY]$, you can't just turn that into $E[X]E[Y]$ unless you know for a fact that $X$ and $Y$ are independent. SymbolizeR is conservative—it won't make that assumption unless you tell it to.

```{r product}
E(X * Y)
```

See? It stays as $E[XY]$. But if you *do* know they are independent, you can say so:

```{r independence}
assume.independent(X, Y)
E(X * Y)  
clear.independence() 
```

Now it factors them out. This explicit control prevents you from making accidental mathematical errors.

---

## Variance and Covariance

We've also taught it the definitions of Variance and Covariance. Under the hood, `Var(X)` basically just expands to $E[X^2] - E[X]^2$. 

This means logic about constants works exactly as you'd expect:

```{r variance}
Var(a * X + b)
```

Because adding a constant $b$ simply shifts the distribution, it doesn't affect the spread, so it disappears from the variance. The scaling factor $a$ comes out squared.

The same logic applies to covariance:

```{r covariance}
Cov(X, Y)
```

---

## Where it Utility Stops (For Now)

It's important to know what SymbolizeR *isn't*. It is not a general-purpose Computer Algebra System like Mathematica. It specializes in **expectations of polynomials**.

If you throw something non-linear at it that it doesn't have a specific rule for, it will often just hand it back to you:

```{r limitations}
E(1/X)    # No general rule for the expectation of an inverse
E(X^Y)    # This is getting too wild
```

It's designed to help you with the messy algebra of moments, derivations, and standard probability manipulations—the stuff that takes up 80% of the time in a standard stats derivation.

---

## Next Steps

Now that you've seen the basics, here's where you can go deeper:

<div class="page-nav">
  <div></div>
  <a href="distributions.html" class="page-nav-item next">
    <span class="page-nav-label">Next</span>
    <span class="page-nav-title">Distributions →</span>
  </a>
</div>

