---
title: "Package Internals"
output: 
  html_document:
    template: template.html
    self_contained: false
    toc: false
    theme: null
    highlight: pygments
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Package Internals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
```

This vignette provides a **technical deep-dive** into SymbolizeR's architecture. It explains how expressions are parsed, transformed, and simplified internally.

```{r setup}
library(SymbolizeR)
```

---

## Architecture Overview

SymbolizeR is built around **R's native expression tree representation**. The core insight is that unevaluated R code (captured via `substitute()` or `quote()`) is a tree structure that we can traverse and transform.

```
User Call: E(a * X + b)
    │
    ▼
┌────────────────────────────┐
│  Expression Tree Capture   │  ← substitute(expr) captures unevaluated AST
└────────────────────────────┘
    │
    ▼
┌────────────────────────────┐
│  Type Classification       │  ← classify.type() distinguishes RV/const/number
└────────────────────────────┘
    │
    ▼
┌────────────────────────────┐
│  Recursive Transformation  │  ← expect.recursive() applies linearity rules
└────────────────────────────┘
    │
    ▼
┌────────────────────────────┐
│  Distribution Lookup       │  ← get.first.moment() substitutes known E[X]
└────────────────────────────┘
    │
    ▼
┌────────────────────────────┐
│  Expression Simplification │  ← simplify.expr() removes identity elements
└────────────────────────────┘
    │
    ▼
Output: a * E(X) + b  (or a * mu + b if X ~ Normal(mu, sigma))
```

---

## File Organization

The R source files are numbered by dependency order:

| File | Purpose |
|------|---------|
| `00_utils.R` | Core utilities: `classify.type`, `ensure.expression`, `expand.poly` |
| `01_internal_expect.R` | Expectation engine: `expect.recursive` |
| `02_internal_simplify.R` | Algebraic simplification: `simplify.expr`, `collect.terms` |
| `03_user_wrappers.R` | Core API: `E()`, `Var()`, `Cov()` |
| `04_higher_order.R` | Higher-order statistics: `Skewness()`, `Kurtosis()` |
| `05_derivations.R` | Step-by-step: `derive.E()`, `derive.Var()`, etc. |
| `06_distributions.R` | Distribution registry: `define()`, `undefine()` |
| `07_independence.R` | Independence: `assume.independent()`, `are.independent()` |
| `08_moments.R` | Moment lookup: `get.first.moment()`, `get.mgf()` |
| `09_higher_moments.R` | 3rd/4th moments: `get.nth.moment()` |
| `10_calculus.R` | Symbolic differentiation: `deriv.sym()` |
| `11_integration.R` | Kernel integration: `integrate.sym()` |
| `12_latex.R` | LaTeX conversion: `to.latex()` |

---

## Expression Trees in R

R expressions are **abstract syntax trees (ASTs)**. Every unevaluated call is a list-like structure:

```{r ast}
# Capture an unevaluated expression
expr <- quote(a * X + b)

# The tree structure
is.call(expr)   # TRUE - it's a call object

# First element is the operator
expr[[1]]       # `+` (the function being called)

# Remaining elements are arguments
expr[[2]]       # a * X (left operand - itself a call)
expr[[3]]       # b (right operand - a symbol)
```

### Tree Visualization

For `a * X + b`, the AST looks like:

```
      (+)
     /   \
   (*)    b
  /   \
 a     X
```

We traverse this tree recursively, transforming nodes according to mathematical rules.

---

## Type Classification (`00_utils.R`)

The `classify.type()` function determines how to treat each node:

| Return Value | Condition | Example |
|--------------|-----------|---------|
| `"number"` | `is.numeric(sym)` | `5`, `2.5` |
| `"rv"` | Symbol starting with uppercase | `X`, `Y`, `Normal` |
| `"const"` | Symbol starting with lowercase | `a`, `mu`, `sigma` |
| `"call"` | Complex expression | `a * X`, `X + Y` |

```{r classify, eval=FALSE}
classify.type(quote(X))       # "rv"
classify.type(quote(mu))      # "const"
classify.type(5)              # "number"
classify.type(quote(a * X))   # "call"
```

This classification drives all downstream behavior: constants pass through unchanged, random variables get wrapped in `E()`, and nested calls recurse.

---

## The Expectation Engine (`01_internal_expect.R`)

The `expect.recursive()` function walks the expression tree, applying:

$$E[aX + b] = aE[X] + b \qquad E[X + Y] = E[X] + E[Y]$$

### Algorithm Pseudocode

```
expect.recursive(expr):
    if expr is numeric:
        return expr
    
    if expr is symbol:
        if lowercase (constant):
            return expr
        if uppercase (RV):
            if has known distribution:
                return first_moment
            else:
                return E(expr)
    
    if expr is call:
        op = operator of expr
        
        if op is '+' or '-':
            return op(expect.recursive(left), expect.recursive(right))
        
        if op is '*':
            classify both operands
            if const * RV:
                return const * expect.recursive(RV)
            if RV * RV:
                if independent:
                    return expect.recursive(left) * expect.recursive(right)
                else:
                    return E(left * right)  # cannot simplify
            ...
```

### Handling Products

The trickiest case is multiplication involving random variables:

```{r products}
# Constant * RV: factor out constant
E(a * X)

# RV * RV without independence: cannot factor
E(X * Y)

# With independence declared:
assume.independent(X, Y)
E(X * Y)
clear.independence()
```

---

## Algebraic Simplification (`02_internal_simplify.R`)

The simplification system is one of the most sophisticated parts of SymbolizeR. It combines multiple techniques to reduce expressions to canonical form.

### Identity Elimination

The first pass removes trivial identity operations:

| Pattern | Simplifies To | Rule |
|---------|---------------|------|
| `x + 0` | `x` | Additive identity |
| `x * 1` | `x` | Multiplicative identity |
| `x * 0` | `0` | Zero property |
| `x^0` | `1` | Zero exponent |
| `x^1` | `x` | Unit exponent |
| `E(constant)` | `constant` | Expectation of constant |

### Like-Term Collection

The key algorithm is **like-term collection**, which combines terms such as:

$$a \cdot X + b \cdot X \to (a+b) \cdot X$$

#### Step 1: Flatten to Terms (`flatten.to.terms`)

Converts nested sums into a flat list of `(coefficient, base)` pairs:

```
Input: 2*X + 3*Y - X
Output: [(2, X), (3, Y), (-1, X)]
```

#### Step 2: Extract Coefficient and Base (`extract.coef.base`)

Separates each term into its numeric coefficient and symbolic base:

```{r coef-base, eval=FALSE}
# Internal structure:
extract.coef.base(quote(3 * X))  # list(coef=3, base=X)
extract.coef.base(quote(X))      # list(coef=1, base=X)
extract.coef.base(quote(-2 * Y)) # list(coef=-2, base=Y)
```

#### Step 3: Canonical Keys (`canonical.key`)

Creates a unique string representation for each base, so we can group like terms:

```{r canonical, eval=FALSE}
canonical.key(quote(X))      # "X^1"
canonical.key(quote(X^2))    # "X^2"
canonical.key(quote(X * Y))  # "X^1*Y^1"  (alphabetically sorted)
```

#### Step 4: Collect and Sum (`collect.terms`)

Groups terms by canonical key, sums coefficients, and reconstructs:

```{r collect, eval=FALSE}
# Before: 2*X + 3*Y - X
# Grouped: {"X^1": [2, -1], "Y^1": [3]}
# Summed:  {"X^1": 1, "Y^1": 3}
# After:   X + 3*Y
```

### Product Normalization (`normalize.product`)

For products like `2 * X * 3 * X^2`, extracts numeric coefficients and combines variable powers:

```
Input:  2 * X * 3 * X^2
Output: list(coef=6, base=X^3)
```

This handles the algebraic identity: $a \cdot x^m \cdot b \cdot x^n = (ab) \cdot x^{m+n}$

### Simplification Pipeline

```
Raw Expression
     │
     ▼
┌──────────────────┐
│ Identity Removal │  ← x+0→x, x*1→x, x^1→x
└──────────────────┘
     │
     ▼
┌──────────────────┐
│ Flatten to Terms │  ← Nested sums → flat list
└──────────────────┘
     │
     ▼
┌──────────────────┐
│ Normalize Powers │  ← x*x → x^2, combine exponents
└──────────────────┘
     │
     ▼
┌──────────────────┐
│ Collect Like     │  ← aX + bX → (a+b)X
└──────────────────┘
     │
     ▼
┌──────────────────┐
│ Reconstruct      │  ← List → expression tree
└──────────────────┘
     │
     ▼
Simplified Expression
```

### Multiple Passes

Some simplifications enable others. For example, after `collect.terms` combines `2*X + 3*X` into `5*X`, identity removal might find new opportunities. The main functions call `simplify.expr` multiple times:

```{r multipass, eval=FALSE}
# In Var():
clean_result <- simplify.expr(simplify.expr(simplify.expr(result)))
```

---

## Polynomial Expansion (`00_utils.R`)

The `expand.poly()` function handles algebraic expansion for variance/covariance calculations:

```{r expand}
# Variance needs (aX + b)^2 expanded
Var(a * X + b)
```

Internally, `expand.poly()` applies:

$$(A + B)^2 = A^2 + 2AB + B^2$$
$$(A \cdot B)^n = A^n \cdot B^n$$

And distributes multiplication over addition:

$$A \cdot (B + C) = A \cdot B + A \cdot C$$

This expansion happens **before** the expectation engine runs, so `E()` sees individual terms rather than parenthesized powers.

---

## Distribution System (`04_distributions.R`)

### The Package Environment

Distributions are stored in a private environment `pkg.env`:

```{r define}
define(X ~ Normal(mu, sigma))
```

This stores:

```{r stored, eval=FALSE}
pkg.env$X = list(
  distribution = "Normal",
  params = list(mu = quote(mu), sigma = quote(sigma))
)
```

### Moment Lookup

When `expect.recursive()` encounters a registered variable, it calls:

- `get.first.moment(var_name)` → Returns symbolic E[X]
- `get.second.moment(var_name)` → Returns symbolic E[X²]
- `get.mgf(var_name, t)` → Returns M_X(t) = E[e^tX]

Each function uses a `switch()` on the distribution name to return the appropriate formula:

```{r moments}
define(X ~ Normal(mu, sigma))
E(X)      # mu (from get.first.moment)
E(X^2)    # sigma^2 + mu^2 (from get.second.moment)
clear.definitions()
```

### Independence Tracking

Independence is stored as pairs in `pkg.env$.independence_pairs`:

```{r independence2}
assume.independent(X, Y, Z)  # Creates pairs: X:Y, X:Z, Y:Z
show.independence()
clear.independence()
```

The `are.independent(x, y)` function checks if a pair exists in this list.

---

## Calculus Module (`05_calculus.R`)

### Symbolic Differentiation

`deriv.sym()` wraps `stats::D()` with special handling for `E()` calls:

```{r deriv}
deriv.sym(x^3, x)           # 3 * x^2
deriv.sym(x * E(Y), x)      # E(Y) - treats E(Y) as constant
```

The `protect.E()` / `restore.E()` pattern replaces `E(...)` with temp symbols `__E_...___` before calling `stats::D()`, then restores them after.

### Definite Integration

`integrate.sym()` uses **kernel recognition** rather than symbolic antiderivatives:

```{r integrate}
# Gamma kernel: x^a * exp(-bx)
integrate.sym(x^2 * exp(-x), x, 0, Inf)

# Gaussian kernel: exp(-ax^2)
integrate.sym(exp(-x^2), x, -Inf, Inf)

# Beta kernel: x^a * (1-x)^b
integrate.sym(x * (1-x), x, 0, 1)
```

If no kernel matches, an unevaluated `Integrate()` call is returned.

---

## LaTeX Conversion (`06_latex.R`)

The `to.latex()` S3 generic dispatches to `expr.to.latex()` for recursive conversion:

```{r latex}
to.latex(E(a * X + b))
to.latex(Var(X))
```

Key conversions:

| R Expression | LaTeX Output |
|--------------|--------------|
| `E(X)` | `\mathbb{E}[X]` |
| `Var(X)` | `\text{Var}(X)` |
| `a/b` | `\frac{a}{b}` |
| `x^2` | `x^{2}` |
| `sqrt(x)` | `\sqrt{x}` |
| `mu`, `sigma` | `\mu`, `\sigma` |

Greek letters are mapped via `.greek_letters`, and subscript notation (e.g., `X_1`) produces proper LaTeX subscripts.

---

## Summary: Data Flow

A complete call like `E(a * X + b)` flows through:

1. **Capture**: `substitute(expr)` → unevaluated AST
2. **Normalize**: `ensure.expression()` → consistent form
3. **Transform**: `expect.recursive()` → apply linearity
4. **Lookup**: `get.first.moment()` → substitute known moments
5. **Simplify**: `simplify.expr()` → remove identities
6. **Return**: Cleaned expression as R call object

The result remains an **unevaluated R expression**—not a string, not a number—which means it can be further manipulated, converted to LaTeX, or used in subsequent calculations.

---



This is where we lift the hood and show you the engine. 

How does SymbolizeR know that `X` is a random variable and `mu` is a constant without you ever telling it? And how does `E(a*X + b)` turn into `a*E(X) + b`?

It uses a technique called **Non-Standard Evaluation (NSE)** to capture your code before it runs, and then it walks through the structure of that code using a **Recursive Descent Parser**.

---

## The Code is a Tree

In R, every expression is actually a tree. When you type `a * X + b`, R sees:

```
    (+)
    / \
  (*)  b
  / \
 a   X
```

SymbolizeR walks down this tree. 

1. It sees the `+`. It knows `E(LHS + RHS) = E(LHS) + E(RHS)`. So it splits the problem in two.
2. It looks at the left branch: `a * X`. It checks if `a` is random (lowercase $\to$ no) and if `X` is random (uppercase $\to$ yes).
3. Since it's `Constant * Random`, it knows it can pull the constant out: `a * E(X)`.
4. It looks at the right branch: `b`. It's a constant. `E(Constant) = Constant`.

The result is reassembled: `a * E(X) + b`.

---

## How it Determines Types

We use a strict but simple heuristic based on variable names. This logic sits in `get.type()`:

1. **Numbers** (like `5` or `3.14`) are always Constants.
2. **Calls** (like `sin(x)` or `exp(y)`) are analyzed recursively. If they contain *any* random variable, the whole call is Random.
3. **Symbols** (variable names) are checked against the convention:
   - Starts with `A-Z` $\to$ **Random Variable**
   - Starts with `a-z` $\to$ **Constant**
   - Starts with `.` $\to$ **Constant**

This is why `mu` works as a constant (starts with lowercase 'm') and `X` works as a variable.

---

## The Expectation Engine

The `E()` function isn't doing any statistics. It's doing **pattern matching**.

Internally, it looks something like this pseudocode:

```r
process_node <- function(node) {
  if (is.sum(node)) {
    return( process_node(node$left) + process_node(node$right) )
  }
  
  if (is.product(node)) {
    if (is.constant(node$left) && is.random(node$right)) {
      return( node$left * E(node$right) )
    }
  }
  
  # ... more rules ...
}
```

It keeps recursing until it hits a base case (like a defined distribution or a generic variable).

---

## Variance is Just Expectation

We didn't write a separate complex engine for `Var()`. We essentially defined it as a macro:

`Var(X)` $\to$ `E(X^2) - (E(X))^2`

When you call `Var(a*X)`, SymbolizeR expands it out:
1. Calculates `E((a*X)^2)` $\to$ `E(a^2 * X^2)` $\to$ `a^2 * E(X^2)`
2. Calculates `(E(a*X))^2` $\to$ `(a * E(X))^2` $\to$ `a^2 * E(X)^2`
3. Subtracts them.

This "rewrite rule" approach ensures that `Var()` is always mathematically consistent with `E()`.

---

## Why did we assume independence sometimes?

You might notice if you run `E(X * Y)`, it doesn't simplify. That's because `E[XY] = E[X]E[Y]` is **only** true if `X` and `Y` are independent (or at least uncorrelated).

We track independence using an internal environment. When you call `assume.independent(X, Y)`, we literally add a tag to the pair `(X, Y)`. The engine checks this tag before applying the product rule.

This makes the system safe. It assumes nothing, unless you or a definition explicitly grants permission.

<div class="page-nav">
  <a href="calculus.html" class="page-nav-item prev">
    <span class="page-nav-label">Previous</span>
    <span class="page-nav-title">← Calculus</span>
  </a>
  <div></div>
</div>
